---
title: "Global COVID-19 Epidemiologic Data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{WHO_SR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(data2019nCoV)

library(tidyr)
library(dplyr)
library(ggplot2)
library(scales)

WHO_SR$Date[length(WHO_SR$Date)]
sarscov2_ecdc_2019$date[length(sarscov2_ecdc_2019$date)]
```

This vignette visualizes data in the R package data2019nCoV, available on [GitHub](https://github.com/eebrown/data2019nCoV). Be sure to reinstall the package to use the latest data, via `devtools::install_github("eebrown/data2019nCoV")`. 

The global data comes from 2 sources, the World Health Organization and the European CDC. 

The data were manually extracted from the WHO [situation reports]( https://www.who.int/emergencies/diseases/novel-coronavirus-2019/situation-reports) and stored in `raw-data/WHO_SR.csv`, which is loaded as the data frame `WHO_SR` in this package. This package uses freely available data, and itself is freely available with the CC0 licence. Information from WHO may be subject to WHO [copyright and terms](https://www.who.int/publishing/copyright/en/). The data source and package should be attributed. No warranty is made as to the accuracy of the data or transcription: use at your own risk. Please file an issue if you find any errors. 

European CDC data was extracted by [Our World in Data](https://ourworldindata.org/coronavirus-source-data).

## Global Cases

```{r, fig.width=6, fig.height=6}

# Global Cases
options(scipen=5)
plot(WHO_SR$Date, WHO_SR$Global.confirmed, 
     main = "Global COVID-19 Cases",
     xlab = "Date",
     ylab = "Confirmed Cases",
     type = "b")

plot(sarscov2_ecdc_2019$date, sarscov2_ecdc_2019$cases_global, 
     main = "Global COVID-19 Cases",
     xlab = "Date",
     ylab = "Confirmed Cases",
     type = "b")
```

### New Daily Cases

```{r, fig.width=6, fig.height=6}
# A function to calculate the daily change
daily_change <- function(series) {
  change <- c(series, NA) - c(NA, series)
  change <- change[-1]
  change <- change[-length(change)]
  return(change)
}

# Remove the change introduced with SR 27, when the definition was expanded.
change_cases <- daily_change(WHO_SR$Global.confirmed)
change_cases_ecdc <- daily_change(sarscov2_ecdc_2019$cases_global)

plot(WHO_SR$Date[-1], change_cases,
     main = "Change in Global Cases by Date",
     ylab = "New Confirmed Cases",
     xlab = "Date",
     type = "b")

plot(sarscov2_ecdc_2019$date[-1], change_cases_ecdc,
     main = "Change in Global Cases by Date",
     ylab = "New Confirmed Cases",
     xlab = "Date",
     type = "b")

```

## Global Cases

```{r, fig.width=7, fig.height=7}

# Every country
gather(sarscov2_ecdc_2019, key, value, 
       
       cases_afg, cases_alb, cases_dza, cases_and, cases_ago, cases_aia, cases_atg, cases_arg, cases_arm, cases_abw, cases_aus, cases_aut, cases_aze, cases_bhs, cases_bhr, cases_bgd, cases_brb, cases_blr, cases_bel, cases_blz, cases_ben, cases_bmu, cases_btn, cases_bol, cases_bes, cases_bih, cases_bwa, cases_bra, cases_vgb, cases_brn, cases_bgr, cases_bfa, cases_bdi, cases_khm, cases_cmr, cases_can, cases_cpv, cases_cym, cases_caf, cases_tcd, cases_chl, cases_chn, cases_col, cases_cog, cases_cri, cases_civ, cases_hrv, cases_cub, cases_cuw, cases_cyp, cases_cze, cases_cod, cases_dnk, cases_dji, cases_dma, cases_dom, cases_ecu, cases_egy, cases_slv, cases_gnq, cases_eri, cases_est, cases_eth, cases_fro, cases_flk, cases_fji, cases_fin, cases_fra, cases_pyf, cases_gab, cases_gmb, cases_geo, cases_deu, cases_gha, cases_gib, cases_grc, cases_grl, cases_grd, cases_gum, cases_gtm, cases_ggy, cases_gin, cases_gnb, cases_guy, cases_hti, cases_hnd, cases_hun, cases_isl, cases_ind, cases_idn, cases_international, cases_irn, cases_irq, cases_irl, cases_imn, cases_isr, cases_ita, cases_jam, cases_jpn, cases_jey, cases_jor, cases_kaz, cases_ken, cases_kosovo, cases_kwt, cases_kgz, cases_lao, cases_lva, cases_lbn, cases_lbr, cases_lby, cases_lie, cases_ltu, cases_lux, cases_mkd, cases_mdg, cases_mwi, cases_mys, cases_mdv, cases_mli, cases_mlt, cases_mrt, cases_mus, cases_mex, cases_mda, cases_mco, cases_mng, cases_mne, cases_msr, cases_mar, cases_moz, cases_mmr, cases_nam, cases_npl, cases_nld, cases_ncl, cases_nzl, cases_nic, cases_ner, cases_nga, cases_mnp, cases_nor, cases_omn, cases_pak, cases_xwb, cases_pan, cases_png, cases_pry, cases_per, cases_phl, cases_pol, cases_prt, cases_pri, cases_qat, cases_rou, cases_rus, cases_rwa, cases_kna, cases_lca, cases_vct, cases_smr, cases_stp, cases_sau, cases_sen, cases_srb, cases_syc, cases_sle, cases_sgp, cases_sxm, cases_svk, cases_svn, cases_som, cases_zaf, cases_kor, cases_ssd, cases_esp, cases_lka, cases_sdn, cases_sur, cases_swz, cases_swe, cases_che, cases_syr, cases_twn, cases_tza, cases_tha, cases_tls, cases_tgo, cases_tto, cases_tun, cases_tur, cases_tca, cases_uga, cases_ukr, cases_are, cases_gbr, cases_usa, cases_vir, cases_ury, cases_uzb, cases_vat, cases_ven, cases_vnm, cases_yem, cases_zmb, cases_zwe) %>%
  ggplot(aes(x=date, y=value, col=key)) +
  geom_line() +
  theme(legend.position="none") +
  labs(title = "Confirmed Cases by Region",
       x = "Date", 
       y = "Confirmed Cases") +
  theme(legend.title = element_blank())
```


# Chinese Provinces

```{r, fig.width=7, fig.height=7}
gather(WHO_SR, key, value, China.Hubei,
       China.Guangdong, China.Beijing, China.Shanghai, China.Chongqing,        
       China.Zhejiang, China.Jiangxi, China.Sichuan, China.Tianjin, China.Henan,
       China.Hunan, China.Shandong, China.Yunnan, China.Taiwan, China.Taipei, 
       China.HongKongSAR, China.Macao, China.Unspecified, China.Anhui, China.Jiangsu,   
       China.Fujian, China.Shaanxi, China.Guangxi, China.Hebei, China.Heilongjiang,
       China.Liaoning, China.Hainan, China.Shanxi, China.Gansu, China.Guizhou, China.Ningxia,
       China.InnerMongolia, China.Xinjiang, China.Jilin, China.Qinghai, China.Xizang) %>%
  ggplot(aes(x=Date, y=value, col=key)) +
  geom_line() +
  scale_y_continuous(trans = 'log10', labels = comma) +
  labs(title = "Cases by Chinese Province (Semilogarithmic)",
       x = "Date", 
       y = "Confirmed Cases") +
  theme(legend.title = element_blank())

```

## Major Outbreaks

```{r, fig.width=6, fig.height=6}
# Major Outbreaks

gather(sarscov2_ecdc_2019, key, value, 
       cases_chn, 
       cases_ita,
       cases_irn,
       cases_kor,
       cases_esp, 
       cases_deu, 
       cases_fra,  
       cases_usa,
       cases_che, #Switzerland, 
       cases_nld, #Netherlands, 
       cases_gbr, #UnitedKingdom, 
       cases_aut, #Austria
       cases_tur,
       ) %>%
  ggplot(aes(x=date, y=value, col=key)) +
  geom_line() +
  theme(legend.position="right", legend.title = element_blank()) +
  scale_y_continuous(trans = 'log10', labels = comma) +
  labs(title = "Major COVID-19 Outbreaks (Semilogarithmic)",
       x = "Date", 
       y = "Confirmed Cases") 

overx <- function(country, x) {
  over <- country[country > x] 
  over <- c(over, rep(NA, length(WHO_SR$Date)))
  return(over)
}

start_no <- 5000

plot(overx(WHO_SR$China, start_no), col="gray", type="l",
     xlab=paste0("Days Since ", start_no, " Cases"),
     ylab= "Cases",
     main=paste0("Outbreak Progression from ", start_no, " Cases"),
     xlim=c(0,60),
     ylim=c(1000,700000))
lines(overx(WHO_SR$Italy, start_no), col="green")
lines(overx(WHO_SR$Iran, start_no), col="orange")
lines(overx(WHO_SR$RepublicofKorea, start_no), col="magenta")
lines(overx(WHO_SR$Spain, start_no), col="brown")
lines(overx(WHO_SR$Germany, start_no), col="coral")
lines(overx(WHO_SR$France,  start_no), col="purple")
lines(overx(WHO_SR$UnitedStatesofAmerica, start_no), col="blue")
lines(overx(WHO_SR$Switzerland,  start_no), col="black")
lines(overx(WHO_SR$Austria,  start_no), col="black")
lines(overx(WHO_SR$UnitedKingdom,  start_no), col="cornflowerblue")
lines(overx(WHO_SR$Netherlands, start_no), col="black")
lines(overx(WHO_SR$Norway, start_no), col="black")
lines(overx(WHO_SR$Sweden, start_no), col="black")
lines(overx(WHO_SR$Denmark, start_no), col="black")
lines(overx(WHO_SR$Belgium, start_no), col="black")
lines(overx(WHO_SR$Canada, start_no), col="red")

legend(x = "right", legend = c("USA", "Spain", "Germany", 
                               "Italy", "China", "UK", 
                               "France", "Iran", "Korea", "Canada"),
       col =    c("blue", "brown",   "coral",  
                  "green", "gray", "cornflowerblue", 
                  "purple", "orange", "magenta", "red"), 
       pch=18)

```

### Outbreaks - Daily Cases

```{r, fig.width=6, fig.height=20}
outbreaks <- list(China = WHO_SR$China, 
                  Italy = WHO_SR$Italy,
                  Canada = WHO_SR$Canada,
                  Spain = WHO_SR$Spain,
                  Germany = WHO_SR$Germany,
                  Iran = WHO_SR$Iran,
                  France = WHO_SR$France,
                  Korea = WHO_SR$RepublicofKorea)

par(mfrow=c(4,2))

for (i in 1:length(outbreaks)) {
  
 plot(WHO_SR$Date[-1], daily_change(outbreaks[[i]]),
      main = names(outbreaks)[[i]],
      ylab = "Daily Cases",
      xlab = "Date",
      type = "l",
      ylim = c(0,10000))
}

 par(mfrow=c(4,2))
 
 plot(WHO_SR$Date[-1], daily_change(WHO_SR$UnitedStatesofAmerica),
      main = "United States - Cases",
      ylab = "Daily Cases",
      xlab = "Date",
      type = "l",
      ylim = c(0,40000))
 
  plot(WHO_SR$Date[-1], daily_change(WHO_SR$UnitedStatesofAmerica.deaths),
      main = "United States - Deaths",
      ylab = "Daily Cases",
      xlab = "Date",
      type = "l",
      ylim = c(0,4000))
  
   plot(WHO_SR$Date[-1], daily_change(WHO_SR$Italy),
      main = "Italy - Cases",
      ylab = "Daily Cases",
      xlab = "Date",
      type = "l",
      ylim = c(0,40000))
   
    plot(WHO_SR$Date[-1], daily_change(WHO_SR$Italy.deaths),
      main = "Italy - Deaths",
      ylab = "Daily Cases",
      xlab = "Date",
      type = "l",
      ylim = c(0,4000))
 
     plot(WHO_SR$Date[-1], daily_change(WHO_SR$Spain),
      main = "Spain - Cases",
      ylab = "Daily Cases",
      xlab = "Date",
      type = "l",
      ylim = c(0,40000))
     
      plot(WHO_SR$Date[-1], daily_change(WHO_SR$Spain.deaths),
      main = "Spain - Deaths",
      ylab = "Daily Cases",
      xlab = "Date",
      type = "l",
      ylim = c(0,4000))
      
           plot(WHO_SR$Date[-1], daily_change(WHO_SR$France),
      main = "France - Cases",
      ylab = "Daily Cases",
      xlab = "Date",
      type = "l",
      ylim = c(0,40000))
     
      plot(WHO_SR$Date[-1], daily_change(WHO_SR$France.deaths),
      main = "France - Deaths",
      ylab = "Daily Cases",
      xlab = "Date",
      type = "l",
      ylim = c(0,4000))

```

## Deaths

```{r, fig.width=6, fig.height=6}
plot(WHO_SR$Date, WHO_SR$China.deaths,
     main = "Cumulative COVID-19 Deaths in China",
     xlab = "Date",
     ylab = "Deaths",
     type = "b")

# Change in Cases Between Reports
plot(WHO_SR$Date[-1], daily_change(WHO_SR$China.deaths),
     main = "Change in Deaths in China by Date",
     ylab = "New Deaths",
     xlab = "Date",
     type = "b")

Deaths.nonChina <- c(WHO_SR$Deaths.nonChina[1:56], 
                      (WHO_SR$Region.WesternPacific.deaths[57:length(WHO_SR$Date)] + 
                         WHO_SR$Region.European.deaths[57:length(WHO_SR$Date)] +
                         WHO_SR$Region.SouthEastAsia.deaths[57:length(WHO_SR$Date)] +
                         WHO_SR$Region.EasternMediterranean.deaths[57:length(WHO_SR$Date)] +
                         WHO_SR$Region.Americas.deaths[57:length(WHO_SR$Date)] +
                         WHO_SR$Region.African.deaths[57:length(WHO_SR$Date)] +
                         WHO_SR$Region.Other.deaths[57:length(WHO_SR$Date)] -
                         WHO_SR$China.deaths[57:length(WHO_SR$Date)]))

plot(WHO_SR$Date, 
     Deaths.nonChina,
     main = "Cumulative COVID-19 Deaths Outside of China",
     xlab = "Date",
     ylab = "Deaths",
     type = "b")

plot(WHO_SR$Date[-1], daily_change(Deaths.nonChina),
     main = "Change in Deaths Outside China by Date",
     xlab = "Date",
     ylab = "New Deaths",
     type = "b")
```

### Case Fatality Rate

```{r, fig.width=6, fig.height=6}

matplot(as.Date(WHO_SR$Date), cbind( 
                           (WHO_SR$China.deaths/ WHO_SR$China)*100,
                           (WHO_SR$RepublicofKorea.deaths / WHO_SR$RepublicofKorea)*100,
                           (WHO_SR$Italy.deaths / WHO_SR$Italy)*100,
                           (WHO_SR$UnitedStatesofAmerica.deaths / 
                                WHO_SR$UnitedStatesofAmerica)*100,
                           (WHO_SR$Canada.deaths / WHO_SR$Canada)*100,
                           (WHO_SR$Spain.deaths / WHO_SR$Spain)*100),
     main = "Case Fatality Rate",
     xlab = "Date",
     ylab = "Deaths / Confirmed Cases (%)",
     type = "l",
     lty = "solid",
     col = c("black",   "blue",  "green", "magenta", "red", "brown"),
     ylim = c(0,14),
     xaxt="n")
dates<-format(WHO_SR$Date,"%b-%d")
axis(1, at=WHO_SR$Date, labels=dates)
legend(x="top", 
       legend = c("China", "Korea", "Italy", "USA", "Canada", "Spain"), 
       col =    c("black",   "blue",  "green", "magenta", "red", "brown"), 
       pch=18)

```


```{r, fig.width=6, fig.height=6}
population_US <- 329968629
population_CAN <- 37894799

range <- 60:length(sarscov2_ecdc_2019$cases_can)

matplot(as.Date(sarscov2_ecdc_2019$date[range]), cbind( 
                           ( (sarscov2_ecdc_2019$cases_can / population_CAN)[range] * 100000 ),
                           ( (sarscov2_ecdc_2019$deaths_can / population_CAN)[range] * 1000000 ),
                           ( (sarscov2_ecdc_2019$cases_usa / population_US)[range] * 100000 ),
                           ( (sarscov2_ecdc_2019$deaths_usa / population_US)[range] * 1000000 )
                           ),
                           
     main = "Cases and Deaths Per Capita",
     xlab = "Date (2020)",
     ylab = "Cases Per 100k; Deaths per Million",
     type = "l",
     col = c("red",   "red",  "blue", "blue"),
     lty = c("solid", "dotted", "solid", "dotted"),
     ylim = c(0,300),
     ylog = TRUE,
     xaxt="n")
dates<-format(sarscov2_ecdc_2019$date,"%b %d")
axis(1, at=sarscov2_ecdc_2019$date, labels=dates)
legend(x="top", 
       legend = c("Canada Cases per 100k", "Canada Deaths per Million", "USA Cases per 100k", "USA Deaths per Million", "Estimates of Annual Influenza Deaths per Million"), 
       col =    c("red",   "red",  "blue", "blue", "grey"),
       lty = c("solid", "dotted", "solid", "dotted", "dotted"),
       pch=18)
#annual mortality from influenza per million (https://ipac-canada.org/influenza-resources.php)
abline(h = 20, col="grey", lty="dotted") 
#lower and upper annual mortality from influenza per million (CDC)
abline(h = 36, col="grey", lty="dotted") 
abline(h = 184, col="grey", lty="dotted") 
# (https://www.ontario.ca/page/flu-facts)
abline(h = 92, col="grey", lty="dotted") 

```