---
title: "WHO Situation Report Data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{WHO_SR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(data2019nCoV)

library(tidyr)
library(dplyr)
library(ggplot2)
```

This vignette briefly demonstrates the use of data from the R package data2019nCoV, available on [GitHub](https://github.com/eebrown/data2019nCoV). Be sure to reinstall the package to use the latest data, via `devtools::install_github("eebrown/data2019nCoV")`. 

This data was manually extracted from the WHO [situation reports]( https://www.who.int/emergencies/diseases/novel-coronavirus-2019/situation-reports) and stored in `raw-data/WHO_SR.csv`, which is loaded as the data frame `WHO_SR` in this package.

## Plot example

```{r, fig.width=6, fig.height=6}
plot(WHO_SR$Date, WHO_SR$Global.confirmed, 
     main = "Global 2019-CoV Confirmed Cases",
     xlab = "Date",
     ylab = "Confirmed Cases",
     type = "b")

plot(WHO_SR$Date, WHO_SR$Global.confirmed, 
     main = "Global 2019-CoV Confirmed Cases with Quadratic Fit",
     xlab = "Date",
     ylab = "Confirmed Cases",
     type = "b")
date_numeric <- as.numeric(WHO_SR$Date - as.Date("2020-01-20"))
date_numeric2 <- date_numeric^2
quad <- lm(WHO_SR$Global.confirmed ~ date_numeric + date_numeric2)
quad_predict <- predict(quad, list(date_numeric, date_numeric2))
lines(WHO_SR$Date, quad_predict, col="green")

summary(quad)

change_cases <- c(WHO_SR$Global.confirmed, NA) - c(NA, WHO_SR$Global.confirmed)
change_cases <- change_cases[-1]
change_cases <- change_cases[-length(change_cases)]
plot(WHO_SR$Date[-1], change_cases,
     main = "Change in Global Cases by Date",
     ylab = "New Confirmed Cases",
     xlab = "Date",
     type = "b")

gather(WHO_SR, key, value, Japan, RepublicofKorea, VietNam, Singapore, Australia, Malaysia, 
       Cambodia, Philippines, Thailand, Nepal, SriLanka, India, UnitedStatesofAmerica, 
       Canada, Finland, France, Germany, Italy, RussianFederation, Spain, Sweden, 
       UnitedKingdom, Belgium, UnitedArabEmirates, InternationalConveyance) %>%
  ggplot(aes(x=Date, y=value, col=key)) +
  geom_line() +
  theme(legend.position="bottom") +
  labs(title = "Confirmed Cases Outside of China",
       x = "Date", 
       y = "Confirmed Cases") +
  theme(legend.title = element_blank())
```

### Log examples

```{r, fig.width=6, fig.height=6}
plot(WHO_SR$Date, WHO_SR$China.deaths,
     main = "2019-CoV Deaths in China (Logarithmic)",
     xlab = "Date",
     ylab = "Deaths",
     log = "y",
     type = "b")

plot(WHO_SR$Date, WHO_SR$Global.confirmed, 
     main = "Global 2019-CoV Confirmed Cases (Logarithmic)",
     xlab = "Date",
     ylab = "Confirmed Cases",
     log = "y",
     type = "b")

```


