---
title: "WHO Situation Report Data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{WHO_SR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(data2019nCoV)

library(tidyr)
library(dplyr)
library(ggplot2)
```

This vignette briefly demonstrates the use of data from the R package data2019nCoV, available on [GitHub](https://github.com/eebrown/data2019nCoV). Be sure to reinstall the package to use the latest data, via `devtools::install_github("eebrown/data2019nCoV")`. 

This data was manually extracted from the WHO [situation reports]( https://www.who.int/emergencies/diseases/novel-coronavirus-2019/situation-reports) and stored in `raw-data/WHO_SR.csv`, which is loaded as the data frame `WHO_SR` in this package. This package uses freely available data, and itself is freely available with the CC0 licence. Information from WHO may be subject to WHO [copyright and terms](https://www.who.int/publishing/copyright/en/). The data source and package should be attributed. No warranty is made as to the accuracy of the data or transcription: use at your own risk. Please file an issue if you find any errors. 

## Global Cases

```{r, fig.width=6, fig.height=6}

# Global Cases
plot(WHO_SR$Date, WHO_SR$Global.confirmed, 
     main = "Global 2019-CoV Confirmed Cases",
     xlab = "Date",
     ylab = "Confirmed Cases",
     type = "b")

# Change in Cases Between Reports
change_cases <- c(WHO_SR$Global.confirmed, NA) - c(NA, WHO_SR$Global.confirmed)
change_cases <- change_cases[-1]
change_cases <- change_cases[-length(change_cases)]

# Remove the change introduced with SR 27, when the definition was expanded.
change_cases[27] <- NA

plot(WHO_SR$Date[-1], change_cases,
     main = "Change in Global Cases by Date",
     ylab = "New Confirmed Cases",
     xlab = "Date",
     type = "b")

plot(WHO_SR$Date, WHO_SR$Cases.nonChina,
     main = "SARS-CoV-2 Confirmed Cases Excluding China",
     xlab = "Date",
     ylab = "Confirmed Cases",
     type = "b")

# Change in Cases Between Reports - Excluding China
change_cases_nochn <- c(WHO_SR$Cases.nonChina, NA) - c(NA, WHO_SR$Cases.nonChina)
change_cases_nochn <- change_cases_nochn[-1]
change_cases_nochn <- change_cases_nochn[-length(change_cases_nochn)]

plot(WHO_SR$Date[-1], change_cases_nochn,
     main = "Change in Global Cases, Excluding China, by Date",
     ylab = "New Confirmed Cases",
     xlab = "Date",
     type = "b")

```

## Cases by Region

```{r, fig.width=7, fig.height=7}
gather(WHO_SR, key, value, 
       
       RepublicofKorea, Japan, Singapore, Australia, Malaysia, VietNam, 
       Philippines, Cambodia, NewZealand,
       
       Thailand, India, Nepal, SriLanka, 
       
       UnitedStatesofAmerica, Canada, Brazil, 
       
       Italy, Germany, France, UnitedKingdom, Spain, Croatia, Austria, 
       Finland, Israel, RussianFederation, Sweden, Belgium, Denmark, 
       Estonia, Georgia, Greece, NorthMacedonia, Norway, Romania, Switzerland, 
       Belarus, Lithuania, Netherlands,
       
       Iran, Kuwait, Bahrain, UnitedArabEmirates, Iraq, Oman, Lebanon, Pakistan,
       Afghanistan, Egypt,
       
       Algeria, Nigeria,
       
       InternationalConveyance) %>%
  ggplot(aes(x=Date, y=value, col=key)) +
  geom_line() +
  theme(legend.position="bottom") +
  labs(title = "Confirmed Cases Outside of China",
       x = "Date", 
       y = "Confirmed Cases") +
  theme(legend.title = element_blank())

# Western Pacific and Southeast Asia
gather(WHO_SR, key, value, 
       RepublicofKorea, Japan, Singapore, Australia, Malaysia, VietNam, 
       Philippines, Cambodia, NewZealand,
       
       Thailand, India, Nepal, SriLanka) %>%
  ggplot(aes(x=Date, y=value, col=key)) +
  geom_line() +
  theme(legend.position="bottom") +
  labs(title = "Western Pacific and Southeast Asia (Excluding China)",
       x = "Date", 
       y = "Confirmed Cases") +
  theme(legend.title = element_blank())

# Americas, Europe
gather(WHO_SR, key, value, 
       UnitedStatesofAmerica, Canada, Brazil, 
       
       Italy, Germany, France, UnitedKingdom, Spain, Croatia, Austria, 
       Finland, Israel, RussianFederation, Sweden, Belgium, Denmark, 
       Estonia, Georgia, Greece, NorthMacedonia, Norway, Romania, Switzerland, 
       Belarus, Lithuania, Netherlands
       
       ) %>%
  ggplot(aes(x=Date, y=value, col=key)) +
  geom_line() +
  theme(legend.position="bottom") +
  labs(title = "Americas and Europe",
       x = "Date", 
       y = "Confirmed Cases") +
  theme(legend.title = element_blank())

# Eastern Mediterranean and Africa
gather(WHO_SR, key, value, 
       
       Iran, Kuwait, Bahrain, UnitedArabEmirates, Iraq, Oman, Lebanon, Pakistan,
       Afghanistan, Egypt,
       
       Algeria, Nigeria
       
       ) %>%
  ggplot(aes(x=Date, y=value, col=key)) +
  geom_line() +
  theme(legend.position="bottom") +
  labs(title = "Eastern Mediterranean and Africa",
       x = "Date", 
       y = "Confirmed Cases") +
  theme(legend.title = element_blank())

# Outbreaks Outside of China
gather(WHO_SR, key, value, 
       Japan, Singapore, InternationalConveyance, RepublicofKorea, Iran, Italy) %>%
  ggplot(aes(x=Date, y=value, col=key)) +
  geom_line() +
  theme(legend.position="bottom") +
  labs(title = "Outbreaks Outside of China",
       x = "Date", 
       y = "Confirmed Cases") +
  theme(legend.title = element_blank())


# Chinese Provinces
gather(WHO_SR, key, value, 
       China.Guangdong, China.Beijing, China.Shanghai, China.Chongqing,        
       China.Zhejiang, China.Jiangxi, China.Sichuan, China.Tianjin, China.Henan,
       China.Hunan, China.Shandong, China.Yunnan, China.Taiwan, China.Taipei, 
       China.HongKongSAR, China.Macao, China.Unspecified, China.Anhui, China.Jiangsu,   
       China.Fujian, China.Shaanxi, China.Guangxi, China.Hebei, China.Heilongjiang,
       China.Liaoning, China.Hainan, China.Shanxi, China.Gansu, China.Guizhou, China.Ningxia,
       China.InnerMongolia, China.Xinjiang, China.Jilin, China.Qinghai, China.Xizang) %>%
  ggplot(aes(x=Date, y=value, col=key)) +
  geom_line() +
  theme(legend.position="bottom") +
  labs(title = "Cases by Chinese Province, excluding Hubei",
       x = "Date", 
       y = "Confirmed Cases") +
  theme(legend.title = element_blank())

```

## Deaths

```{r, fig.width=6, fig.height=6}
plot(WHO_SR$Date, WHO_SR$China.deaths,
     main = "2019-CoV Deaths in China",
     xlab = "Date",
     ylab = "Deaths",
     type = "b")

plot(WHO_SR$Date, WHO_SR$China.deaths,
     main = "2019-CoV Deaths in China (Logarithmic)",
     xlab = "Date",
     ylab = "Deaths",
     log = "y",
     type = "b")

plot(WHO_SR$Date, (WHO_SR$China.deaths / WHO_SR$China)*100,
     main = "Deaths / Cases in China",
     xlab = "Date",
     ylab = "Deaths / Confirmed Cases (%)",
     type = "b")

plot(WHO_SR$Date, (WHO_SR$Deaths.nonChina / WHO_SR$Cases.nonChina)*100,
     main = "Deaths / Cases (Outside China)",
     xlab = "Date",
     ylab = "Deaths / Confirmed Cases (%)",
     type = "b")

# Change in Cases Between Reports
change_deaths <- c(WHO_SR$China.deaths, NA) - c(NA, WHO_SR$China.deaths)
change_deaths <- change_deaths[-1]
change_deaths <- change_deaths[-length(change_deaths)]
plot(WHO_SR$Date[-1], change_deaths,
     main = "Change in Deaths in China by Date",
     ylab = "New Deaths",
     xlab = "Date",
     type = "b")

plot(WHO_SR$Date, 
     (WHO_SR$France.deaths + WHO_SR$Philippines.deaths + WHO_SR$Japan.deaths +
        WHO_SR$InternationalConveyance.deaths + WHO_SR$RepublicofKorea.deaths +
        WHO_SR$Iran.deaths),
     main = "SARS-CoV-2 Deaths outside of China",
     xlab = "Date",
     ylab = "Deaths",
     type = "b")

```
